Evaluation of probabilistic photometric redshift estimation approaches for \textsc{LSST}

### 2019-03-15 paper/tex/analysis.tex
cutting cruft

The goal of this study is to evaluate the degree to which \pzpdf s of each method can be trusted for a generic probabilistic data analysis.
Obviously \pzpdf s must not take negative values and must integrate to unity over the range of possible redshifts.
The overloaded ``$p(z)$'' is a widespread abuse of notation that obfuscates this goal, so we will dedicate some attention to breaking it apart.
Galaxies have redshifts $z$ and photometric data $d$ drawn from a joint probability space $p(z, d)$ in nature.
As a result, each observed galaxy $i$ has a \textit{true posterior \pzpdf}\ $p(z \vert d_{i})$ as well as a true likelihood $p(d \vert z_{i})$.
There are a number of metrics that can be used to test the accuracy of a \pz\ posterior as an estimator of a true \pz\ posterior if the true \pzpdf\ is known.
However, the true \pzpdf\ is in general not accessible unless the photometry is in fact drawn from the true \pz\ likelihoods, a mock catalogue generation procedure that has not yet appeared in the literature and was not performed for this work.

@@

The exact implementation of the metrics theoretically depends on the parametrization of the \pzpdf s, which may differ across codes and can affect the precision of the estimator \citep{Malz:qp}.
Even considering a single method under the same parametrization, such as the 200-bin $0 < z < 2$ piecewise constant function used here, the exact bin definitions will affect the result.
%\red{Note: Let's address how the gridding (endpoints and spacing) affects the metrics, from the email correspondence Re: [LSST-DESC-MEMBERS] Summary of Nov 29 PZ Telecon. (AIM)}
The piecewise constant format is chosen because of its established presence in the literature, and the choice of 200 bins was motivated by the approximate number of columns expected to be available for storage of \pzpdf s for the final \lsst\ Project tables.\footnote{See, e.~g.~the \lsst\ Data Products Definition Document, available at: \url{https://ls.st/dpdd}}
We will discuss the choice of \pzpdf\ parameterization further in Section~\ref{sec:results}.

This analysis is conducted using the \qp\footnote{\url{http://github.com/aimalz/qp/}}\ software package \citep{Malz:qp} for manipulating and calculating metrics of univariate PDFs.
We present the metrics of \pzpdf s that address our goals in the sections below.
Though the outmoded practice should not be encouraged, those seeking a connection to previous comparison studies will find metrics of common redshift point estimators derived from \pzpdf s in Appendix~\ref{sec:pointmetrics}.

\subsection{Metric of individual \pzpdf s}
\label{sec:CDE_loss}
%(Alex Malz, Ann Lee)

@@

A quantile of a distribution is the value $q$ at which the CDF of the distribution is equal to $Q$; percentiles and quartiles are familiar examples of linearly spaced sets of 100 and 4 quantiles, respectively.
The quantile-quantile (QQ) plot serves as a graphical visualization for comparing two distributions, where the quantiles of one distribution are plotted against the quantiles of the other distribution,
%\blue{Include an equation to ensure quantile definition is crystal clear before building upon it for PIT.}\red{Is maybe the above text clear enough?  Text may be more clear than any equation I can come up with--SJS.}
providing an easy way to qualitatively assess the consistency between an estimating distribution and a true distribution.
%location, scale and skewness.
The closer the QQ plot is to diagonal, the closer the match between the distributions.
% In this paper, QQ plots are used for two purposes: (1) for comparing $N(z)$ from \pzpdf s (estimated using Eq.~\ref{eq:stacked}) with the true $N(z)$, and (2) for assessing the overall consistency of an ensemble of \pzpdf s with their true redshifts on a population level, where the distribution of the PIT values (see previous section) is compared to a uniform distribution between $0$ and $1$.
%\red{is this enough reason to include both?--SJS}
%\aim{Maybe we should explain why we expect the distribution of CDFs to be uniform.}\red{Note: this is just describing the PIT.  I'm going to swap the order of PIT to come before this, and then say we are evaluating the distribution of PIT values.  I will also add why we expect this to be uniform--SJS}

The probability integral transform (PIT)
\begin{align}
\label{eq:pit}
\mathrm{PIT} &\equiv \rm{CDF}[\hat{p}, z_{\rm true}] %\int_{-\infty}^{z_{\rm true}} p(z) dz
\end{align}
is the CDF of a \pzpdf\ evaluated at its true redshift, and the distribution of PIT values probes the average accuracy of the \pzpdf s of an ensemble of galaxies.
The distribution of PIT values is effectively the derivative of the QQ plot.
A catalogue of accurate \pzpdf s should have a PIT distribution that is uniform $U(0,1)$, and deviations from flatness are interpretable: overly broad \pzpdf s induce underrepresentation of the lowest and highest PIT values, whereas overly narrow \pzpdf s induce overrepresentation of the lowest and highest PIT values.
Catastrophic outliers with a true redshift outside the support of its \pzpdf\ have $\mathrm{PIT} \approx 0$ or $\mathrm{PIT} \approx 1$.

@@

% \subsection{Metrics over estimated probability distributions}
% \label{sec:quantmet}

% We compute metrics comparing the CDF of PIT values to the CDF of a Uniform distribution, and also compute the CDF of the true redshift distribution $N'(z)$ compared the $\hat{N}(z)$ distribution derived from summing the \pzpdf s as described in Eq.~\ref{eq:stacked}.

@@

% \subsubsection{Kolmogorov-Smirnov (KS) and related statistics}
% \label{sec:ks}

We evaluate a number of quantitative metrics derived from the visually interpretable QQ plot and PIT histogram, built on the Kolmogorov-Smirnov (KS) statistic
\begin{equation}
  \label{eq:ks}
  \rm{KS} \equiv \max_{z} \left( \left| \rm{CDF}[\hat{f}, z] - \rm{CDF}[\tilde{f}, z] \right| \right),
\end{equation}
interpretable as the maximum difference between the CDFs of an approximating univariate distribution $\hat{f}(z)$ and a reference distribution $\tilde{f}(z)$.
% $CDF_{\rm phot}(z)$ and $CDF_{\rm spec}(z)$, the CDFs of the photo-$z$ and spectroscopic redshift respectively:
%The KS test quantifies whether the two redshift distributions are drawn from a similar distribution, independent of binning.
% The KS test quantifies the similarity between two distributions, independent of binning.
% A lower $N_{\rm KS}$ value corresponds to more similar distributions.
%We calculate two KS test statistics, one to test the distribution of $p(z)$ values, and one to test the stacked estimator $\hat{n}(z)$ of the redshift distribution.
%The $p(z)$ test compares the CDF of the PIT values described in the previous section to that of a uniform distribution.
%The $n(z)$ test compares the CDF of the distribution formed by stacking the $p(z)$ for every object in our test set against the CDF of the corresponding true redshifts.
%In this case a lower $N_{\rm KS}$ value means that the photo-$z$ distribution fits the spectroscopic redshift distribution well. $F_{\rm phot}(z)$ is calculated by stacking the PDFs of every single object in the sample, producing a smooth cumulative $n(z)$ to be compared with $F_{\rm spec}(z)$.
We also consider two variants of the KS statistic.
A cousin of the KS statistic, the Cramer-von Mises (CvM) statistic
\begin{equation}
\label{eq:cvm}
  \rm{CvM}^{2} \equiv \int_{-\infty}^{+\infty} \big(\rm{CDF}[\hat{f}, z] - \rm{CDF}[\tilde{f}, z]\big)^2 \mathrm{d}\rm{CDF}[\tilde{f}, z]
\end{equation}
is the mean-squared difference between the CDFs of an approximate and true PDF.
The Anderson-Darling (AD) statistic
\begin{equation} \label{eq:ad}
  \rm{AD}^2 \equiv N_{tot}\int_{-\infty}^{+\infty} \frac{\big(\rm{CDF}[\hat{f}, z] - \rm{CDF}[\tilde{f}, z]\big)^2} {\rm{CDF}[\tilde{f}, z] (1 -\rm{CDF}[\tilde{f}, z]} \mathrm{d}\rm{CDF}[\tilde{f}, z]
\end{equation}
is a weighted mean-squared difference featuring enhanced sensitivity to discrepancies in the tails of the distribution.
% The CvM statistic is similar to the KS statistic as it is also computed from the distance between the measured CDF and the ideal CDF, but instead of the maximum distance, the CvM statistic
% is the average of the distance squared.
% In this paper we use the CvM statistic to evaluate the performance of both $n(z)$ and $p(z)$.
% The AD statistic
% is a weighted version of the CvM statistic, making it more sensitive to the tails of the distribution, where $N_{tot}$ is the sample size.
%The KS statistic and the CvM and AD statistics related to it are all used to evaluate the performance of both $\hat{n}(z)$ and the QQ and PIT of the CDFs of the ensemble of photo-$z$ interim posteriors.
In anticipation of a substantial fraction of galaxies having PIT of 0 or 1, a consequence of catastrophic outliers, we evaluate the AD statistic with modified bounds of integration $(0.01, 0.99)$ to exclude those extremes in the name of numerical stability.

%\subsubsection{Kullback-Leibler Divergence (KLD)}
%\label{sec:kl}
%(Alex Malz)
%
%The KLD is a measure of the loss in information when using a probability distribution that is an estimate of a known true probability distribution, so we use
%\begin{equation}
%\mathrm{KLD} = \int_{-\infty}^{\infty}n'(z)\log\left[\frac{n'(z)}{\hat{n}(z)}\right]dz
%\end{equation}
%as a metric on the stacked redshift distribution estimator of Eq.~\ref{eq:stacked}. Noting the assumptions made in using Eq.~\ref{eq:stacked}, we say that a photo-$z$ PDFs method whose stacked redshift distribution estimator $\hat{n}(z)$ more closely resembles the true redshift distribution $n'(z)$ is desired, i.e. we seek to minimize the KLD.

>>>>

The goal of this study is to evaluate the degree to which \pzpdf s of each method can be trusted for a generic analysis.
The overloaded ``$p(z)$'' is a widespread abuse of notation that obfuscates this goal, so we will dedicate some attention to breaking it apart.
Galaxies have redshifts $z$ and photometric data $d$ drawn from a joint probability space $p(z, d)$ in nature.
As a result, each observed galaxy $i$ has a \textit{true posterior \pzpdf}\ $p(z \vert d_{i})$ as well as a true likelihood $p(d \vert z_{i})$.
There are a number of metrics that can be used to test the accuracy of a \pz\ posterior as an estimator of a true \pz\ posterior if the true \pzpdf\ is known.
However, the true \pzpdf\ is in general not accessible unless the photometry is in fact drawn from the true \pz\ likelihoods, a mock catalogue generation procedure that has not yet appeared in the literature and was not performed for this work.

@@

The exact implementation of the metrics theoretically depends on the parametrization of the \pzpdf s, which may differ across codes and can affect the precision of the estimator \citep{Malz:qp}.
Even considering a single method under the same parametrization, such as the 200-bin $0 < z < 2$ piecewise constant function used here, the exact bin definitions will affect the result.
%\red{Note: Let's address how the gridding (endpoints and spacing) affects the metrics, from the email correspondence Re: [LSST-DESC-MEMBERS] Summary of Nov 29 PZ Telecon. (AIM)}
The piecewise constant format is chosen because of its established presence in the literature, and the choice of 200 bins was motivated by the approximate number of columns expected to be available for storage of \pzpdf s for the final \lsst\ Project tables.\footnote{See, e.~g.~the \lsst\ Data Products Definition Document, available at: \url{https://ls.st/dpdd}}
We will discuss the choice of \pzpdf\ parameterization further in Section~\ref{sec:discussion}.

This analysis is conducted using the \qp\footnote{\url{http://github.com/aimalz/qp/}}\ software package \citep{Malz:qp} for manipulating and calculating metrics of univariate PDFs.
We present the metrics of \pzpdf s that address our goals in the sections below.
Section~\ref{sec:qualmet} outlines aggregate metrics of a catalogue of \pzpdf s, and Section~\ref{sec:CDE_loss} presents a metric of individual \pzpdf s in the absence of true \pzpdf s.
Though the outmoded practices should not be encouraged, those seeking a connection to previous comparison studies will find metrics of redshift point estimate reductions of \pzpdf s in Appendix~\ref{sec:pointmetrics} and metrics of a science-specific summary statistics heuristically derived from \pzpdf s in Appendix~\ref{sec:moments}.

@@

A quantile of a distribution is the value $q$ at which the CDF of the distribution is equal to $Q$; percentiles and quartiles are familiar examples of linearly spaced sets of 100 and 4 quantiles, respectively.
The quantile-quantile (QQ) plot serves as a graphical visualization for comparing two distributions, where the quantiles of one distribution are plotted against the quantiles of the other distribution, providing an easy way to qualitatively assess the consistency between an estimating distribution and a true distribution.
The closer the QQ plot is to diagonal, the closer the match between the distributions.
% In this paper, QQ plots are used for two purposes: (1) for comparing $N(z)$ from \pzpdf s (estimated using Eq.~\ref{eq:stacked}) with the true $N(z)$, and (2) for assessing the overall consistency of an ensemble of \pzpdf s with their true redshifts on a population level, where the distribution of the PIT values (see previous section) is compared to a uniform distribution between $0$ and $1$.
%\red{is this enough reason to include both?--SJS}
%\aim{Maybe we should explain why we expect the distribution of CDFs to be uniform.}\red{Note: this is just describing the PIT.  I'm going to swap the order of PIT to come before this, and then say we are evaluating the distribution of PIT values.  I will also add why we expect this to be uniform--SJS}

The probability integral transform (PIT)
\begin{align}
\label{eq:pit}
\mathrm{PIT} &\equiv \rm{CDF}[\hat{p}, z_{\rm true}]
\end{align}
is the CDF of a \pzpdf\ evaluated at its true redshift, and the distribution of PIT values probes the average accuracy of the \pzpdf s of an ensemble of galaxies.
The distribution of PIT values is effectively the derivative of the QQ plot.
A catalogue of accurate \pzpdf s should have a PIT distribution that is uniform $U(0,1)$, and deviations from flatness are interpretable: overly broad \pzpdf s induce underrepresentation of the lowest and highest PIT values, whereas overly narrow \pzpdf s induce overrepresentation of the lowest and highest PIT values.
Catastrophic outliers with a true redshift outside the support of its \pzpdf\ have $\mathrm{PIT} \approx 0$ or $\mathrm{PIT} \approx 1$.

@@

@@

We evaluate a number of quantitative metrics derived from the visually interpretable QQ plot and PIT histogram, built on the Kolmogorov-Smirnov (KS) statistic
\begin{equation}
  \label{eq:ks}
  \rm{KS} \equiv \max_{z} \left( \left| \rm{CDF}[\hat{f}, z] - \rm{CDF}[\tilde{f}, z] \right| \right),
\end{equation}
interpretable as the maximum difference between the CDFs of an approximating univariate distribution $\hat{f}(z)$ and a reference distribution $\tilde{f}(z)$.
% $CDF_{\rm phot}(z)$ and $CDF_{\rm spec}(z)$, the CDFs of the photo-$z$ and spectroscopic redshift respectively:
%The KS test quantifies whether the two redshift distributions are drawn from a similar distribution, independent of binning.
% The KS test quantifies the similarity between two distributions, independent of binning.
% A lower $N_{\rm KS}$ value corresponds to more similar distributions.
%We calculate two KS test statistics, one to test the distribution of $p(z)$ values, and one to test the stacked estimator $\hat{n}(z)$ of the redshift distribution.
%The $p(z)$ test compares the CDF of the PIT values described in the previous section to that of a uniform distribution.
%The $n(z)$ test compares the CDF of the distribution formed by stacking the $p(z)$ for every object in our test set against the CDF of the corresponding true redshifts.
%In this case a lower $N_{\rm KS}$ value means that the photo-$z$ distribution fits the spectroscopic redshift distribution well. $F_{\rm phot}(z)$ is calculated by stacking the PDFs of every single object in the sample, producing a smooth cumulative $n(z)$ to be compared with $F_{\rm spec}(z)$.
We also consider two variants of the KS statistic.
A cousin of the KS statistic, the Cramer-von Mises (CvM) statistic
\begin{equation}
\label{eq:cvm}
  \rm{CvM}^{2} \equiv \int_{-\infty}^{+\infty} \big(\rm{CDF}[\hat{f}, z] - \rm{CDF}[\tilde{f}, z]\big)^2 \mathrm{d}\rm{CDF}[\tilde{f}, z]
\end{equation}
is the mean-squared difference between the CDFs of an approximate and true PDF.
The Anderson-Darling (AD) statistic
\begin{equation} \label{eq:ad}
  \rm{AD}^2 \equiv N_{tot}\int_{-\infty}^{+\infty} \frac{\big(\rm{CDF}[\hat{f}, z] - \rm{CDF}[\tilde{f}, z]\big)^2} {\rm{CDF}[\tilde{f}, z] (1 -\rm{CDF}[\tilde{f}, z]} \mathrm{d}\rm{CDF}[\tilde{f}, z]
\end{equation}
is a weighted mean-squared difference featuring enhanced sensitivity to discrepancies in the tails of the distribution.
%The KS statistic and the CvM and AD statistics related to it are all used to evaluate the performance of both $\hat{n}(z)$ and the QQ and PIT of the CDFs of the ensemble of photo-$z$ interim posteriors.
In anticipation of a substantial fraction of galaxies having PIT of 0 or 1, a consequence of catastrophic outliers, we evaluate the AD statistic with modified bounds of integration $(0.01, 0.99)$ to exclude those extremes in the name of numerical stability.

%\subsubsection{Kullback-Leibler Divergence (KLD)}
%\label{sec:kl}
%(Alex Malz)
%The KLD is a measure of the loss in information when using a probability distribution that is an estimate of a known true probability distribution, so we use
%\begin{equation}
%\mathrm{KLD} = \int_{-\infty}^{\infty}n'(z)\log\left[\frac{n'(z)}{\hat{n}(z)}\right]dz
%\end{equation}
%as a metric on the stacked redshift distribution estimator of Eq.~\ref{eq:stacked}. Noting the assumptions made in using Eq.~\ref{eq:stacked}, we say that a photo-$z$ PDFs method whose stacked redshift distribution estimator $\hat{n}(z)$ more closely resembles the true redshift distribution $n'(z)$ is desired, i.e. we seek to minimize the KLD.

\subsection{Conditional Density Estimate (CDE) Loss: a metric of individual \pzpdf s}
\label{sec:CDE_loss}
%(Alex Malz, Ann Lee)
